---
title: "Reproducible Report"
author:
  - name: 
      given: Anna
      family: Shattuck
    affiliation: VectorByte 
citation: true
date: 2026-05-29
format: html
editor: visual
---

```{r, echo=FALSE, include = F, message = F}

# if DL = true, download from vecdyn but if false,, use saved r files
DL = FALSE

# Load libraries we need
library(tidyverse)
library(terra)
library(sf)
```

```{r, include = F, message = F}
# here i can load datasets if i have already downlaoded them and saved them in the WD

if(DL == FALSE){
  load("data_raw.Rsave")
  load("mosq_site.Rsave")
}
```

```{r, include = F, message = F, eval=DL}
# We can hide our set up code chunk using the "echo = FALSE" option. 
# We can make sure no messages pop up by adding "message = FALSE".
# Include = FALSE doesn't show the code OR the output

# Get the API ready to access data
source("code/VecDyn_Dataset_Access.R")


# Get data sets
data_raw <- getDataset(679)

save(data_raw, file = "code/data_raw.Rsave") 
```

```{r, echo=FALSE, eval = DL}
######## Data Cleaning ##########

# # Collapse the datasets into one dataset
# mosqDF <- data.frame()
# 
# for (i in 1:length(list_of_dataframes)) {
#   
#   df <- list_of_dataframes[[i]]
#   
#   mosqDF <- rbind(mosqDF, df)
#   
# }

# filter raw data for albopictus and aegypti
mosq <- data_raw %>% filter(genus == "Aedes", 
                            species %in% c("albopictus", "aegypti"))

# Make sure date is a date variable
mosq$sample_start_date <- ymd(mosq$sample_start_date)
mosq$sample_end_date <- ymd(mosq$sample_end_date)

# create a site index and join to mosquito data based on unique long/lat coords
sites <- mosq %>% 
  distinct(sample_lat_dd, sample_long_dd)

sites <- sites %>% mutate(site_index = seq(1:nrow(sites)))

# join this back to mosquito data
mosq_site <- full_join(mosq, sites, by = c("sample_lat_dd","sample_long_dd"))


# # Combine observations from the same date
# mosqDFcomb <- mosqDF %>% 
#   group_by(sample_start_date) %>%
#   summarize(totalMos = sum(sample_value))

# Add year and month variables
mosq_site$month <- month(mosq_site$sample_start_date)
mosq_site$year <- year(mosq_site$sample_start_date)

save(mosq_site, file = "code/mosq_site.Rsave") 
```


## Background

This report includes *Aedes albopictus* and *Ae. aegypti* abundance data collected by the Orange County Florida Mosquito Control group using CDC light traps. These data are from `{r} length(unique(mosq_site$year))` and span months `{r} min(mosq_site$month, na.rm = TRUE)` to `{r} max(mosq_site$month, na.rm = TRUE)`.

The dataset contains a total of `{r} mosq_site %>% filter(species == "albopictus") %>% summarise(total = sum(sample_value, na.rm = TRUE)) %>% pull(total)` *Ae. albopictus* collected and `{r} mosq_site %>% filter(species == "aegypti") %>% summarise(total = sum(sample_value, na.rm = TRUE)) %>% pull(total)` *Ae. aegypti* collected.

A total of `{r} mosq_site %>% distinct(site_index) %>% nrow()` sites were sampled across the year. 

Some field sites collected only one of the species, both, or neither. This county houses the city of Orlando, FL.

```{r, echo=FALSE, message=FALSE}
# create a plot of all 72 sites
boundary_AEA<-st_read("OrangeCountyShp/OrangeCountyShp.shp",quiet=T)%>%
  st_transform(crs=crs("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))%>%
  dplyr::select(geometry)%>%
  st_union()%>%
  as_Spatial()
boundary_AEA <- st_as_sf(boundary_AEA)

mosq_year <- mosq_site %>% 
  group_by(site_index, sample_long_dd, sample_lat_dd) %>% 
  summarise(
    caught_albo = any(species == "albopictus" & sample_value > 0, na.rm = TRUE),
    caught_aeg = any(species == "aegypti" & sample_value > 0, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(
    species_group = case_when(
      caught_albo & caught_aeg  ~ "Both species",
      caught_albo & !caught_aeg ~ "albopictus only",
      !caught_albo & caught_aeg ~ "aegypti only",
      TRUE                 ~ "Neither species"
    )
  )

ggplot()+
  geom_sf(data=boundary_AEA,aes(geometry=geometry), fill="grey90",color="black")+
  geom_point(data=mosq_year,
              aes(x = sample_long_dd, y = sample_lat_dd, color = species_group), size = 1.5) +
  labs(x = "Longitude", y = "Latitude", color = "Species Presence")+
  theme_minimal() + 
  theme(legend.position = c(.9, .9))+
  theme(legend.position = "right",
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 10),
        axis.title.x = element_text(size = 15),
        axis.text.y = element_text(size = 10),
        axis.title.y = element_text(size = 15),
        axis.line = element_line(color = "black"),
        axis.ticks = element_line(color = "black"),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 13))
```

Site sampling occurred fairly consistently throughout the year at over half the sites included in the data set. Calendar week two was the only week the had consistent gaps in sampling which would warrant further inquiry if we were to use this data set in the future. 

```{r, echo=FALSE, , message=FALSE}
# plot to explore the number of sites that were sampled across weeks

# create week 
mosq_site$week <- week(mosq_site$sample_start_date)

# # # Combine observations from the same week, species, and site
mosq_week <- mosq_site %>%
  group_by(week, site_index, species) %>%
  summarize(totalMos = sum(sample_value))

mosq_week_binned <- mosq_week %>%
  mutate(
    totalMos_bin = case_when(
      totalMos == 0 ~ "0",
      totalMos >= 1 & totalMos <= 10 ~ "1–10",
      totalMos >= 11 & totalMos <= 20 ~ "11–20",
      totalMos >= 21 & totalMos <= 50 ~ "21–50",
      totalMos > 50 ~ ">50",
      TRUE ~ NA_character_
    ),
    totalMos_bin = factor(
      totalMos_bin,
      levels = c("0", "1–10", "11–20", "21–50", ">50")
    )
  )

ggplot(mosq_week_binned, aes(x = week, y = site_index, fill = totalMos_bin)) +
  geom_tile(color = "white", linewidth = 0.15) +
  facet_wrap(~ species, nrow = 1) +
  scale_x_continuous(
    breaks = seq(
      min(mosq_week_binned$week, na.rm = TRUE),
      max(mosq_week_binned$week, na.rm = TRUE),
      by = 4
    ),
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c(
      "0" = "grey90",
      "1–10" = "#fee8c8",
      "11–20" = "#fdbb84",
      "21–50" = "#e34a33",
      ">50" = "#7f0000"
    ),
    drop = FALSE,
    na.value = "white"
  ) +
  labs(
    x = "Week",
    y = "Site",
    fill = "Mosquitoes caught"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 8),
    axis.title = element_text(size = 14),
    legend.position = "right",
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11),
    plot.margin = margin(10, 20, 10, 10)
  )

```

As every site was not consistently sampled across the entire year, we adjusted mosquito abundance to include sampling effort to explore trends in mosquito populations in the traps across the year. 

```{r, echo=FALSE, message=FALSE}
#explore total mosquito abundance across our sites for albo and aegypti

# create week 
mosq_site$week <- week(mosq_site$sample_start_date)

# # # Combine observations from the same week, species, and site
mosq_week <- mosq_site %>%
  group_by(week, site_index, species) %>%
  summarize(totalMos = sum(sample_value)) %>% 
  ungroup()

mosq_week_total <- mosq_week %>% 
  group_by(week, species) %>% 
  summarise(
    totalMos_w = sum(totalMos, na.rm = TRUE),
    n_traps = n_distinct(site_index),
    mos_per_trap = totalMos_w / n_traps,
    .groups = "drop"
  )

ggplot(mosq_week_total, aes(x = week, y = mos_per_trap, color = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    x = "Week",
    y = "Mosquitoes per trap",
    color = "Species"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank()
  )+
  theme(legend.position = "right",
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 10),
        axis.title.x = element_text(size = 15),
        axis.text.y = element_text(size = 10),
        axis.title.y = element_text(size = 15),
        axis.line = element_line(color = "black"),
        axis.ticks = element_line(color = "black"),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 13))
  

```


```{r}
load("site_imp.Rdata")


```


